Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-03 forecastflows.tex
adding some refs

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong discrepancies exist between model projections \citep{Cheaib2012}. Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches. 
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter identification, and uncertainty quantification. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. However, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.

>>>>

These limitations are central to current debates. In climate modeling, increasing model complexity has not necessarily led to reduced uncertainty. For instance, the uncertainty range on the effect of increasing CO2 concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

@@

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
An universal workflow offers an opportunity to bridge statistical and process-based frameworks, integrating mechanistic knowledge and leveraging robust statistical approaches \citep[e.g.][]{rounce2020quantifying}.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter identification, and uncertainty quantification. Process-based models would no longer be perceived as deterministic black boxes by other researchers but rather as robust statistical frameworks encapsulating both data structure and mechanistic knowledge. This shift would present a significant challenge---as it would likely reveal many issues related to model degeneracies and data limitations before achieving robust inference. However, it would prevent modelers from making biased inferences and unfounded assumptions beyond what the data can support. And it would also be an opportunity to spread the incorporation of mechanistic assumptions beyond the process-based modeling community.
